### 2025-03-02 tutorials/getstarted.qmd
add a final code example for the second section.
First full draft of the tutorial.

In a nutshell a Lie group $𝒢$ is a [manifold](https://en.wikipedia.org/wiki/Manifold) that also is a group a group operation.
A manifold informally be described as a set that “locally looks like a Euclidean space”
and has some way to measure angles at every point, to be precise an inner product, that is allowed to (smoothly)
depend on the point we are at.
A [group](https://en.wikipedia.org/wiki/Group_(mathematics)) means that on the set there also
exists an operation, within this package usually denoted by $∘: 𝒢 × 𝒢 → 𝒢$$,
`LieGroups.jl` uses the manifolds defined in [`Manifolds.jl`](https://juliamanifolds.github.io/Manifolds.jl/stable/). For more details on that, see also the [introductory tutorial](https://juliamanifolds.github.io/Manifolds.jl/stable/tutorials/getstarted/) there.

@@

@@

# TODO
* group action and apply
* product of two Lie groups
* semidirect product

>>>>

In a nutshell a Lie group $𝒢$ is a [manifold](https://en.wikipedia.org/wiki/Manifold) $ℳ$ that also is a group a group operation.
A manifold informally be described as a set that “locally looks like a Euclidean space”
and has some way to measure angles at every point, to be precise an inner product, that is allowed to (smoothly)
depend on the point we are at.
A [group](https://en.wikipedia.org/wiki/Group_(mathematics)) means that on the set there also
exists an operation, within this package usually denoted by $∘: 𝒢 × 𝒢 → 𝒢$$,
`LieGroups.jl` uses the manifolds defined in [`Manifolds.jl`](https://juliamanifolds.github.io/Manifolds.jl/stable/). For more details on that, see also the [introductory tutorial](https://juliamanifolds.github.io/Manifolds.jl/stable/tutorials/getstarted/) there.
A Lie group is usually written as a tuple $(ℳ,∘)$ os the manifold and the group operation thereon.

@@

As an example, we compute

```{julia}
h = exp(SO3, g, X)
```
and its inverse is returning `X` as well.
```{julia}
log(SO3, g, h)
```

@@

There are two further ingredients this tutorial needs to get towards rigid body motions.

### Group actions

For a group like $\mathrm{SO}(3)$ an action describes how points $g ∈ 𝒢$ can act on
points from a Riemannian manifold $p ∈ ℳ$ – in short an action of a certain [`AbstractGroupActionType`](@ref)
”combines“ $p$ and $g$ into a new element $q = σ_g(p)$.
In the example of $𝒢=\mathrm{SO}(3)$ and $ℳ=ℝ^3$ the action is (just) the application of the
rotation or the matrix-vector multiplication $σ_g(p)=gp$.
A special case is, when $ℳ$ itself is also a Lie group.
In the example this is the case, since together with vector addition $p+q$ we get the [`TranslationGroup`](@ref) $T(3) = (ℝ^3, +)$
Depending on how the concatenation of two types, the [`AbstractLeftGroupActionType`](@ref) like the one here
and a [`AbstractRightGroupActionType`](@ref).

```{julia}
T3 = TranslationGroup(3)
```
### Products and semidirect products of Lie groups

Similar to [product manifolds](https://juliamanifolds.github.io/ManifoldsBase.jl/stable/metamanifolds/#ProductManifold)
(direct) [product Lie groups](../groups/product_group.md) are formed the same way. For two Lie groups
$𝒢 = 𝒢_1 × 𝒢_2$ is the product of the two manifolds together with the component wise application of the group operations:
the first group operation (of $𝒢_1$) to the first component and that of $𝒢_2$ to the second.

Instead of the component wise or “not interacting” variant of the (direct) product Lie groups,
for the [semidirect product Lie groups](../groups/semidirect_product_group.md) $𝒢 = 𝒢_1 ⋉ 𝒢_2$
we also require an action of how $𝒢_1$ acts on $𝒢_2$.
Semidirect here means that the first component of $𝒢$ is the same as for the direct product,
but before applying the second group operation on $𝒢_2$ one of the elements is “acted upon”
from an element of $𝒢_1$. The group operation reads for $(g_1,g_2), (h_1,h_2) ∈ 𝒢$

```math
(g_1,g_2) ∘ (h_1,h_2) := (g_1 ⋆ h_1, σ_{h_1}(g_2) ⋄ h_2).
```

### Rigid body motions

We obtain the [special Euclidean group](../groups/special_euclidean_group.md)
$\mathrm{SE}(3) = \mathrm{SO}(3) ⋉ T(3)`` where the group action is the one discussed as an example before.

```{julia}
SE3 = SpecialEuclideanGroup(3)
```

which we could also generate with `SO3` and `T3` from before as

```{julia}
SO3 ⋉ T3
```

This call employs the [`default_left_action`](@ref)`(SO3,T3)` to determine the action
for the semidirect product Lie group. This is defined whenever for two Lie groups, their (left)
action is clear, because there exists a reasonable default. Otherwise the full form

```{julia}
LeftSemidirectProductLieGroup(SO3, T3, LeftGroupOperationAction())
```

is necessary. Here, the first `Left` for the semidirect product refers to the fact that
the left group acts on the right one before the right group operation is performed.
[`LeftGroupOperationAction`](@ref) refers to that the group operation – left multiplication with a matrix – is applied here.

For this case, any point $\mathrm{SE}(3)$ is a tuple of a rotation matrix and a vector.
We model this using [RecursiveArrayTools.jl](https://docs.sciml.ai/RecursiveArrayTools/stable/),
namely its `ArrayPartition`.
For example with the rotation matrix `g` from before we have

```{julia}
g1 = ArrayPartition(g, [1.0, 0.0, 2.0])
```

and for a pure translation we can reuse `e` as in

```{julia}
h1 = ArrayPartition(e, [0.0, 3.0, 0.0])
```

and we can summarize a few operations from before: both are valid points on `SE3`

```{julia}
(is_point(SE3, g1), is_point(SE3, h1))
```

we can perform the group operation

```{julia}
gh1 = compose(SE3, g1, h1)
```

apply the Lie group exponential

```{julia}
Y = log(SE3, gh1)
```

and look at the manifold dimension, it is the sum of the single ones

```{julia}
manifold_dimension(SE3)
```

and its coordinates as

```{julia}
vee(LieAlgebra(SE3), gh1)
```

which actually is component wise again, the first 3 values refer to the rotation part,
the second three to the translation part.
